---
title: "ST2195_Coursework"
author: "Loo Zhi Xian"
date: "2023-03-12"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

##This coursework will be using years 2006 and 2007, as well as airports/carriers/planedata csvs

##Please set working directory to folder where the coursework dataset and dataverse files are located:

### Import the libraries and data

```{r}
library(dplyr)
library(ggplot2)
library(tidyverse)
library(magrittr)
library(moments)
library(hrbrthemes)
```

### Import data from the CSV files

```{r}
airports <- read.csv("airports.csv", header = TRUE)
planes <- read.csv("plane-data.csv", header = TRUE)
carriers <- read.csv("carriers.csv", header = TRUE)
data_2006 <- read.csv("2006.csv")
data_2007 <- read.csv("2007.csv")
```
Row binding the data from 2006 and 2007 and removing duplicates. We also change all empty data into NA values.

```{r}
flight <- rbind(data_2006, data_2007) %>%
  distinct()

flight[flight == ""] <- NA
```

### Question 1 data wrangling

```{r}
#Remove cancelled and delayed flights, then group the flights by their CRS Departure Time
delayed <- flight %>% 
  filter(Cancelled == 0 & Diverted == 0) %>% 
  mutate(DepInterval = cut(CRSDepTime,breaks = c(0, 200, 400, 600, 800, 1000, 1200, 1400, 1600, 1800, 2000, 2200, 2400),
                           labels =  c("0000 ~ 0159 hrs", "0200 ~ 0359 hrs", "0400 ~ 0559 hrs", "0600 ~ 0759 hrs", 
                                       "0800 ~ 0959 hrs", "1000 ~ 1159 hrs", "1200 ~ 1359 hrs", "1400 ~ 1559 hrs",
                                       "1600 ~ 1759 hrs","1800 ~ 1959 hrs", "2000 ~ 2159 hrs", "2200 ~ 2359 hrs"),
                           right = FALSE))

arrival_delay <- delayed %>% 
  filter(ArrDelay > 0) %>%
  select(Year, Month, DayofMonth, DayOfWeek, DepTime, CRSDepTime, ArrTime, CRSArrTime, TailNum, DepDelay, ArrDelay, DepInterval)

arrival_delay
```


```{r}
#mean depart delay time for each intervals
TimeOfDay <- arrival_delay %>%
  group_by(DepInterval) %>%
  summarize(AvgDelay = round(mean(ArrDelay), 2))

TimeOfDay
```


```{r}
#Plot the graph
least_delay_day <-ggplot(TimeOfDay, aes(x= AvgDelay, y=DepInterval)) +
  geom_col(fill= 'pink') +
  labs(title = "Average Delay Duration per Departure Interval", x = "Duration of Delay (Minutes)", y = "") +
  scale_y_discrete(limits = rev) 
least_delay_day + geom_text(aes(label = AvgDelay), hjust = - 0.1 ,position = position_dodge(width = 1), size = 3) +
  theme_classic()

```

```{r}
day_of_week <- arrival_delay%>%
  mutate(DayOfWeek = factor(x=DayOfWeek, 
                            levels = c(1,2,3,4,5,6,7), 
                            labels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))) %>%
  group_by(DayOfWeek) %>%
  summarise(AvgDelay = round(mean(ArrDelay), 2)) %>%
  ggplot(aes(x = AvgDelay, y = DayOfWeek)) +
  geom_col(fill= '#2fccc7') +
  labs(title = "Average Delay Duration Per Day of Week", x = 'Delay Duration (Minutes)', y = '') +
  scale_y_discrete(limits = rev)

day_of_week + geom_text(aes(label = AvgDelay), hjust = - 0.1 ,position = position_dodge(width = 1), size = 3)


```



```{r}
#Quarter
delayed_data_month <- arrival_delay %>% 
  mutate(MonthQuarter = factor(case_when(
    Month  >= 1 & Month <= 3 ~ "Jan - Mar",
    Month >= 4 & Month <= 6 ~ "Apr - Jun",
    Month >= 7 & Month <= 9 ~ "Jul - Sep",
    Month >= 10 & Month <= 12 ~ "Oct - Dec")))

#bar plot for quarterly
quarter_delay <- delayed_data_month %>%
  group_by(MonthQuarter) %>%
  summarise(AvgDelay = round(mean(ArrDelay), 2)) %>%
  ggplot(aes(x = MonthQuarter, y = AvgDelay)) +
  geom_col(fill = '#4bb2cc') +
  scale_x_discrete(limits=rev) +
  labs(title = 'Average Delay Duration for each Quarter', x = "", y = "Delay Duration (Minutes)")

quarter_delay + geom_text(aes(label = AvgDelay), vjust = -0.1, position = position_dodge(width = 1), size = 5)
```

```{r}
#bar plot for month
monthly_delay <- arrival_delay %>%
  mutate(Month = factor(Month)) %>%
  group_by(Month) %>%
  summarise(AvgDelay = round(mean(ArrDelay),2)) %>%
  ggplot(aes(x= AvgDelay, y=Month)) +
  geom_col(fill= '#d15836') +
  scale_y_discrete(limits=rev) +
  labs(title = "Average Delay Duration per Month", x = "Delay Duration (Minutes)", y = "")

monthly_delay + geom_text(aes(label = AvgDelay), hjust = - 0.1, size = 3)
```

##Question 2

```{r}
#join planes to data (duration of delay)
pdata <-left_join(delayed, planes, by = c("TailNum" = "tailnum")) %>%
  rename("ManYear" = "year") %>%
  filter(!(DepDelay <= 0 & ArrDelay <= 0)) #filter out rows with no delays

#change the type from chr to int and drop all rows with NA in 'year' column
pdata %<>% mutate(ManYear = as.integer(ManYear)) %>%
  drop_na(ManYear)

pdata

#show only ArrDelay and year column, filter out flights with no year
ydata <- pdata %>% 
  filter(ArrDelay > 0, ManYear != 0) %>%
  select(ArrDelay, ManYear) %>%
  group_by(ManYear) %>%
  summarise(AvgDelay = mean(ArrDelay))

ydata

```

```{r}
ggplot(ydata, aes(x = ManYear, y = AvgDelay)) +
  geom_point() +
  geom_smooth()+
  labs(title = "Average Delay for Planes", x = "Manufacture Year for Plane", y = "Average Delay (Minutes)") +
  theme_ipsum()
```

### Question 3

```{r}
flight$route <- paste(flight$Origin, flight$Dest, sep = "-")

travel <- flight %>%
  group_by(route, Year) %>%
  summarise(TotalFlights = n())

before <- filter(travel, Year == "2006")
before %<>% rename(Before = TotalFlights)

after <- filter(travel, Year == "2007")
after %<>% rename(After = TotalFlights)

difference <- left_join(before, after, by = "route") %>%
  select(route, Before, After)

difference <- replace(difference,is.na(difference), 0)

h6 <- difference %>%
  summarize(Difference = After - Before) %>%
  arrange(desc(Difference))

head(h6) #increased the most
tail(h6) #decreased the most

```

```{r}
#another method, by incoming and outgoing difference for each year, for each state
info <- left_join(flight, airports, by = c("Origin" = "iata")) %>%
  left_join(airports, by= c("Dest" = "iata"))

info <- info %>%
  rename('DestState' = 'state.y','OrigState' = 'state.x') %>%
  select(Year, Origin, Dest, OrigState, DestState)

outbound <- info %>%
  group_by(OrigState, Year) %>%
  count() %>%
  pivot_wider(names_from = Year, values_from = n) %>%
  mutate(Diff = `2007` - `2006`) %>%
  rename(State = OrigState) %>%
  na.omit() %>%
  arrange(desc(Diff))

head(outbound) #states with the highest increase in outbound flights
tail(outbound) #states with the highest decrease in outbound flights

inbound <- info %>%
  group_by(DestState, Year) %>%
  count() %>%
  pivot_wider(names_from = Year, values_from = n) %>%
  mutate(Diff = `2007` - `2006`) %>%
  rename(State = DestState) %>%
  na.omit() %>%
  arrange(desc(Diff))

head(inbound) #states with the highest increase in inbound flight
tail(inbound) #states with the highest decrease in inbound flight
```

```{r}
ggplot(head(inbound), aes(x= State, y= Diff))  +
  geom_col() +
  labs(title = "Highest increase in inbound flights", x = "States", y = "") +
  theme_light()

ggplot(tail(inbound), aes(x= State, y= Diff))  +
  geom_col() +
  labs(title = "Highest decrease in inbound flights", x = "States", y = "") +
  theme_light()  

ggplot(head(outbound), aes(x= State, y= Diff))  +
  geom_col() +
  labs(title = "Highest increase in outbound flights", x = "States", y = "") +
  theme_light()

ggplot(tail(outbound), aes(x= State, y= Diff))  +
  geom_col() +
  labs(title = "Highest decrease in outbound flights", x = "States", y = "") +
  theme_light()
```
##Question 4
```{r}
#show that departure delay affects arrival delay at the next airport
delayed %>%
  filter(DepDelay > 0) %>%
  select(ArrDelay, DepDelay) %>%
  distinct() %>%
  ggplot(aes(x = DepDelay, y = ArrDelay)) +
  geom_point() +
  geom_smooth(method= lm, color="red", se=FALSE) +
  labs(title = "Relationship between Departure and Arrival Delay", x = "Departure Delay (Minutes)", y = "Arrival Delay (Minutes)") +
  theme_ipsum()

#from delayed, check the airport with the most LateAircraftDelay
delayed %>%
  filter(LateAircraftDelay > 0) %>%
  group_by(Origin) %>%
  summarise(Total = n()) %>%
  arrange(desc(Total)) %>%
  head()

#find the tailnum with the most occurence of late aircraft delay; N679AE
delayed %>%
  filter(Origin == "ORD", LateAircraftDelay > 0) %>%
  group_by(TailNum) %>% 
  summarise(Total = n()) %>%
  arrange(desc(Total)) %>%
  head()

#From the filters, can see on 2 january 2006, the flight from MSP to ORD had a delay, which affected the subsequent flights
fdata <- flight %>%
  filter(TailNum == "N679AE") %>%
  select(Year, Month, DayofMonth, DepTime, CRSDepTime, ArrTime, CRSArrTime, DepDelay, ArrDelay, Origin, Dest) %>%
  arrange(Year,Month, DayofMonth, CRSDepTime) %>%
  filter(Year == "2006", Month == "1", DayofMonth == '2') %>%
  slice(1:3)

fdata

ggplot(fdata)+
  geom_line(aes(y=CRSArrTime,x= Origin,colour="CRSArrTime"),group=1 )+
  geom_point(aes(y=CRSArrTime,x= Origin,colour="CRSArrTime")) +
  geom_line(aes(y=ArrTime,x= Origin,colour="ArrTime"),group=1) +
  geom_point(aes(y=ArrTime,x= Origin,colour="ArrTime"))

```
### Question 5

```{r}
#variables to use: year, month, dayofmonth, dayofweek, 
library(mlr3)
library(mlr3learners)
library(mlr3pipelines)
library(mlr3tuning)
```

Convert to factors if required, and create the column `family_size`:

```{r}
flight %<>%
  mutate(Delayed = ifelse(ArrDelay > 0 | DepDelay > 0, 0, 1 ))

flight$Delayed <- as.factor(flight$Delayed)
flight %<>% na.omit()

```

Get the training and test sets

```{r}
n <- nrow(flight)
train_set <- sample(n, round(0.5*n))
test_set <- setdiff(1:n, train_set)
```

### Set up the task 

```{r}
task <- TaskClassif$new('flight', backend=flight, target = 'Delayed')
task$select(c('Year', 'Month', 'DayofMonth', 'DayOfWeek', 'CRSDepTime'))
measure <- msr('classif.ce')
task
```

Some of the variables are factor for which some methods do not support factor, so we need to convert them to numerical values.
(The following will be used later.)

```{r}
fencoder <- po("encode", method = "treatment",
  affect_columns = selector_type("factor"))
ord_to_int <- po("colapply", applicator = as.integer,
  affect_columns = selector_type("ordered"))
```

Some methods require tuning the hyperparameters. 
(The following will be used later.)

```{r}
tuner <- tnr('grid_search')
terminator <- trm('evals', n_evals = 20)
```

## 2. Extend the results with different classification models

### Logistic regression

```{r}
learner_lr <- lrn("classif.log_reg")
gc_lr <- po('imputemean') %>>%
  po(learner_lr)
glrn_lr <- GraphLearner$new(gc_lr)

glrn_lr$train(task, row_ids = train_set)
glrn_lr$predict(task, row_ids = test_set)$score() 
```

### Gradient boosting

```{r}
install.packages("xgboost")
learner_gb <- lrn("classif.xgboost")
gc_gb <- po('imputemean') %>>%
  fencoder %>>% ord_to_int %>>%
  po(learner_gb)
glrn_gb <- GraphLearner$new(gc_gb)

glrn_gb$train(task, row_ids = train_set)
```

```{r}
glrn_gb$predict(task, row_ids = test_set)$score() 
```

### Penalised logistic regression

```{r echo = TRUE, results = 'hide'}
learner_plr <- lrn('classif.glmnet') 
gc_plr <- po('scale') %>>% 
  fencoder %>>% ord_to_int %>>%
  po('imputemean') %>>%
  po(learner_plr)
glrn_plr <- GraphLearner$new(gc_plr)
tune_lambda <- ParamSet$new (list(
 ParamDbl$new('classif.glmnet.lambda', lower = 0.001, upper = 2)
))

at_plr <- AutoTuner$new(
  learner = glrn_plr,
  resampling = rsmp('cv', folds = 3),
  measure = measure,
  search_space = tune_lambda,
  terminator = terminator,
  tuner = tuner
)
at_plr$train(task, row_ids = train_set)
```

```{r}
at_plr$predict(task, row_ids = test_set)$score()
```

### Classification trees

```{r}
learner_tree <- lrn("classif.rpart")

gc_tree <- po('imputemean') %>>%
  po(learner_tree)
glrn_tree <- GraphLearner$new(gc_tree)

glrn_tree$train(task, row_ids = train_set)
glrn_tree$predict(task, row_ids = test_set)$score() 
```

### Random forests

```{r echo = TRUE, results = 'hide'}
learner_rf <- lrn('classif.ranger') 
learner_rf$param_set$values <- list(min.node.size = 4)
gc_rf <- po('scale') %>>%
  po('imputemean') %>>%
  po(learner_rf)
glrn_rf <- GraphLearner$new(gc_rf)
tune_ntrees <- ParamSet$new (list(
 ParamInt$new('classif.ranger.num.trees', lower = 50, upper = 600)
))
at_rf <- AutoTuner$new(
  learner = glrn_rf,
  resampling = rsmp('cv', folds = 3),
  measure = measure,
  search_space = tune_ntrees,
  terminator = terminator,
  tuner = tuner
)
at_rf$train(task, row_ids = train_set)
```

```{r}
at_rf$predict(task, row_ids = test_set)$score() 
```

### Support vector machines

```{r}
install.packages("e1071")
learner_svm <- lrn("classif.svm")

gc_svm <- po('imputemean') %>>% 
  fencoder %>>% ord_to_int %>>%
  po(learner_svm)
glrn_svm <- GraphLearner$new(gc_svm)

glrn_svm$train(task, row_ids = train_set)
glrn_svm$predict(task, row_ids = test_set)$score() 
```

```{r results = 'hide'}
set.seed(1) # for reproducible results
# list of learners
lrn_list <- list(
  glrn_lr,
  glrn_gb,
  at_plr,
  glrn_tree,
  at_rf,
  glrn_svm
)
# set the benchmark design and run the comparisons
bm_design <- benchmark_grid(task = task, resamplings = rsmp('cv', folds = 3), learners = lrn_list)
bmr <- benchmark(bm_design, store_models = TRUE)
```

```{r}
library(mlr3viz)
library(ggplot2)
autoplot(bmr) + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
bmr$aggregate(measure)
```