---
title: "ST2195_Coursework"
author: "Loo Zhi Xian"
date: "2023-03-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

##This coursework will be using years 2006 and 2007, as well as airports/carriers/planedata csvs

##Please set working directory to folder where the coursework dataset and dataverse files are located:

### Import the libraries and data

```{r}
library(dplyr)
library(ggplot2)
library(tidyverse)
library(magrittr)
library(moments)
library(hrbrthemes)
```

### Import data from the CSV files

```{r}
airports <- read.csv("airports.csv", header = TRUE)
planes <- read.csv("plane-data.csv", header = TRUE)
data_2006 <- read.csv("2006.csv")
data_2007 <- read.csv("2007.csv")
```

Join the data from 2006 and 2007 together and remove duplicates.
```{r}
flight <- rbind(data_2006, data_2007) %>%
  distinct()

#Change all empty data into NA values.
flight[flight == ""] <- NA
```

### Question 1 data wrangling

```{r}
#Remove cancelled and delayed flights, then group the flights by their CRS Departure Time
flight %<>%
  filter(Cancelled == 0 & Diverted == 0) %>% 
  mutate(DepInterval = cut(CRSDepTime,breaks = c(0, 200, 400, 600, 800, 1000, 1200, 1400, 1600, 1800, 2000, 2200, 2400),
                           labels =  c("0000 ~ 0159 hrs", "0200 ~ 0359 hrs", "0400 ~ 0559 hrs", "0600 ~ 0759 hrs", 
                                       "0800 ~ 0959 hrs", "1000 ~ 1159 hrs", "1200 ~ 1359 hrs", "1400 ~ 1559 hrs",
                                       "1600 ~ 1759 hrs","1800 ~ 1959 hrs", "2000 ~ 2159 hrs", "2200 ~ 2359 hrs"),
                           right = FALSE)) %>%
  mutate(ADelay = as_factor(ifelse(ArrDelay > 0, 1,0))) %>%
  mutate(DDelay = as_factor(ifelse(DepDelay > 0, 1,0))) %>%
  select(-TaxiIn, -TaxiOut, -CancellationCode, -Diverted, -Cancelled, -UniqueCarrier)

arrival_delay <- flight %>% 
  filter(ArrDelay > 0) %>%
  select(Year, Month, DayofMonth, DayOfWeek, CRSDepTime, DepTime, DepDelay, ArrDelay, DepInterval)

arrival_delay
```
##Best time of the day

```{r}
#Probabiliy a flight being delayed for each depature interval
flight %>%
  group_by(DepInterval) %>% 
  summarize(Percent = round(mean(ADelay == 1) * 100, 2)) %>%
  ggplot(aes(y= Percent, x= DepInterval)) +
  geom_col(fill='steelblue') +
  scale_y_continuous(limits= c(0, 60), expand = c(0, 0)) +
  labs(title = "Percentage of delayed flights per interval", x = "", y = "Percentage (%)") +
  scale_x_discrete() +
  geom_text(aes(label=Percent), position=position_dodge(width=0.9), vjust=-0.25) +
  theme_classic() +
  theme(axis.text.x = element_text(face="bold", angle = 45, hjust=1))


#mean depart delay time for each intervals
TimeOfDay <- arrival_delay %>%
  group_by(DepInterval) %>%
  summarize(AvgDelay = round(mean(ArrDelay), 2))

TimeOfDay
```


```{r}
#Plot the graph
least_delay_day <-ggplot(TimeOfDay, aes(x= AvgDelay, y=DepInterval)) +
  geom_col(fill= 'lightblue') +
  labs(title = "Average delay duration per interval", x = "Duration of Delay (Minutes)", y = "") +
  scale_y_discrete(limits = rev)+
  scale_x_continuous(limits= c(0, 45), expand = c(0, 0))
least_delay_day + geom_text(aes(label = AvgDelay), hjust = - 0.1 ,position = position_dodge(width = 1), size = 3) +
  theme_classic()

```

```{r}
flight %>%
  mutate(DayOfWeek = factor(x=DayOfWeek, 
                            levels = c(1,2,3,4,5,6,7), 
                            labels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))) %>%
  group_by(DayOfWeek) %>% 
  summarize(Percent = round(mean(ADelay == 1) * 100, 2)) %>%
  ggplot(aes(x= DayOfWeek, y= Percent)) +
  geom_col(fill= 'steelblue') +
  labs(title = "Percentage of delayed flights per day", x = "Day", y = "Percentage (%)") +
  scale_x_discrete() +
  scale_y_continuous(limits= c(0, 60), expand = c(0, 0)) +
  geom_text(aes(label=Percent), position=position_dodge(width=0.9), vjust=-0.25) +
  theme_classic()

day_of_week <- arrival_delay %>%
  mutate(DayOfWeek = factor(x=DayOfWeek, 
                            levels = c(1,2,3,4,5,6,7), 
                            labels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))) %>%
  group_by(DayOfWeek) %>%
  summarise(AvgDelay = round(mean(ArrDelay), 2)) %>%
  ggplot(aes(x = AvgDelay, y = DayOfWeek)) +
  geom_col(fill= 'lightblue') +
  labs(title = "Average Delay Duration Per day", x = 'Delay Duration (Minutes)', y = '') +
  scale_y_discrete(limits = rev) +
  scale_x_continuous(limits= c(0, 40), expand = c(0, 0)) +
  theme_classic()

day_of_week + geom_text(aes(label = AvgDelay), hjust = - 0.1 ,position = position_dodge(width = 1), size = 3)


```

```{r}
#probability of a flight having an arrival delay
flight %>%
  mutate(Month = factor(Month)) %>%
  group_by(Month) %>% 
  summarize(Percent = round(mean(ADelay == 1) * 100, 2)) %>%
  ggplot(aes(x= Month, y= Percent)) +
  geom_col(fill='steelblue') +
  labs(title = "Probability of delayed flight", x = "Month", y = "Percentage (%)") +
  scale_x_discrete(labels = c('Jan', 'Feb', 'Mar', 'Apr', 'May',
                              'Jun','Jul','Aug','Sep', 'Oct', 'Nov', 'Dec')) +
  scale_y_continuous(limits=c(0,60),expand= c(0,0)) +
  geom_text(aes(label=Percent), position=position_dodge(width=0.9), vjust=-0.25) +
  theme_classic()

#bar plot for month
monthly_delay <- arrival_delay %>%
  mutate(Month = factor(Month)) %>%
  group_by(Month) %>%
  summarise(AvgDelay = round(mean(ArrDelay),2)) %>%
  ggplot(aes(x= AvgDelay, y=Month)) +
  geom_col(fill= 'lightblue') +
  scale_y_discrete(labels = c('Jan', 'Feb', 'Mar', 'Apr', 'May',
                              'Jun','Jul','Aug','Sep', 'Oct', 'Nov', 'Dec'), limits = rev) +
  scale_x_continuous(limits=c(0,40), expand=c(0,0)) +
  labs(title = "Average Delay Duration per Month", x = "Delay Duration (Minutes)", y = "")

monthly_delay + geom_text(aes(label = AvgDelay), hjust = - 0.1, size = 3)
```

##Question 2

```{r, echo=FALSE,warning=FALSE,message=FALSE,error=FALSE, results='hide',fig.keep='all'}
#join planes to flight and change the type from chr to int and drop all rows with NA in 'year' column
planes %<>% select(tailnum, year)

pdata <-left_join(flight, planes, by = c("TailNum" = "tailnum")) %>%
  rename("ManYear" = "year") %>%
  mutate(ManYear = as.integer(ManYear)) %>%
  filter(ManYear != 0 & !is.na(ManYear))
  
#probability of a flight with delay
pdata %>%
  group_by(ManYear) %>%
  summarise(Percent = mean(ADelay == 1 | DDelay == 1)* 100) %>%
  ggplot(aes(x = ManYear, y = Percent)) +
  geom_point() +
  labs(title = "Percentage of flight delay", x = "Manufacture Year", y = "Percentage (%)") +
  theme_classic()
```

### Question 3

```{r}
#by incoming and outgoing difference for each year, for each state
info <- left_join(flight, airports, by = c("Origin" = "iata")) %>%
  left_join(airports, by= c("Dest" = "iata"))

info <- info %>%
  rename('DestState' = 'state.y','OrigState' = 'state.x') %>%
  select(Year, Origin, Dest, OrigState, DestState)

outbound <- info %>%
  group_by(OrigState, Year) %>%
  count() %>%
  pivot_wider(names_from = Year, values_from = n) %>%
  mutate(Diff = `2007` - `2006`) %>%
  rename(State = OrigState) %>%
  na.omit() %>%
  arrange(desc(Diff))

head(outbound) #states with the highest increase in outbound flights
tail(outbound) #states with the highest decrease in outbound flights

inbound <- info %>%
  group_by(DestState, Year) %>%
  count() %>%
  pivot_wider(names_from = Year, values_from = n) %>%
  mutate(Diff = `2007` - `2006`) %>%
  rename(State = DestState) %>%
  na.omit() %>%
  arrange(desc(Diff))

head(inbound) #states with the highest increase in inbound flight
tail(inbound) #states with the highest decrease in inbound flight
```

```{r}
states <- c("AL" = "Alabama", "AK" = "Alaska",  "AZ" = "Arizona",  "AR" = "Arkansas",  "CA" = "California",  "CO" = "Colorado",
  "CT" = "Connecticut",  "DE" = "Delaware",  "FL" = "Florida", "GA" = "Georgia",  "HI" = "Hawaii",  "ID" = "Idaho",  "IL" = "Illinois",
  "IN" = "Indiana",  "IA" = "Iowa",  "KS" = "Kansas",  "KY" = "Kentucky",  "LA" = "Louisiana",  "ME" = "Maine",  "MD" = "Maryland",
  "MA" = "Massachusetts",  "MI" = "Michigan",  "MN" = "Minnesota",  "MS" = "Mississippi",  "MO" = "Missouri",  "MT" = "Montana",
  "NE" = "Nebraska",  "NV" = "Nevada",  "NH" = "New Hampshire",  "NJ" = "New Jersey",  "NM" = "New Mexico",  "NY" = "New York",  "NC" = "North Carolina",
  "ND" = "North Dakota",  "OH" = "Ohio",  "OK" = "Oklahoma",  "OR" = "Oregon",  "PA" = "Pennsylvania",
  "RI" = "Rhode Island",  "SC" = "South Carolina", "SD" = "South Dakota",  "TN" = "Tennessee", "TX" = "Texas",  
  "UT" = "Utah", "VT" = "Vermont",  "VA" = "Virginia", "WA" = "Washington", "WV" = "West Virginia",
  "WI" = "Wisconsin", "WY" = "Wyoming", "PR" = "Puerto Rico", "VI" = "Virgin Islands")

mostout <- head(outbound) %>% 
  rename('Outbound'='Diff')

mostin <- head(inbound)%>%
  rename('Inbound' = 'Diff')

leastout <- tail(outbound) %>%
  rename('Outbound' = 'Diff')

leastin <-tail(inbound)%>%
  rename('Inbound' = 'Diff')

#most increase in inbound and outbound
increase <- rbind(mostout, mostin) %>%
  gather('Direction', 'Value', 'Outbound':'Inbound', na.rm=TRUE)
increase$State <- as.factor(increase$State)

ggplot(increase, aes(x=State, y= Value, fill= Direction)) +
  geom_bar(position= "dodge", stat = "identity") +
  scale_x_discrete(labels = states) +
  scale_y_continuous(limits=c(0, 50000), expand= c(0,0)) +
  labs(title = "Most increase in flights between 2006 and 2007", y= "Difference") +
  theme_light()

#most decrease in inbound and outbound
decrease <-rbind(leastout, leastin) %>%
  gather("Direction", "Value", 'Outbound':'Inbound', na.rm = TRUE)
decrease$State <- as.factor(decrease$State)

ggplot(decrease, aes(x=State, y= abs(Value), fill= Direction)) +
  geom_bar(position= "dodge", stat = "identity") +
  scale_x_discrete(labels = states) +
  scale_y_continuous(limits=c(0, 30000), expand= c(0,0)) +
  labs(title = "Most decrease in flights between 2006 and 2007", y= "Difference") +
  theme_light()
```
##Question 4

```{r, message=FALSE}

#show that departure delay affects arrival delay at the next airport
flight %>%
  filter(DepDelay > 0) %>%
  select(ArrDelay, DepDelay) %>%
  distinct() %>%
  ggplot(aes(x = DepDelay, y = ArrDelay)) +
  geom_point() +
  geom_smooth(method= lm, color="red", se=FALSE) +
  labs(title = "Relationship between Departure and Arrival Delay", x = "Departure Delay (Minutes)", y = "Arrival Delay (Minutes)") +
  theme_classic()

#from delayed, check the airport with the most LateAircraftDelay
flight %>%
  filter(LateAircraftDelay > 0) %>%
  group_by(Origin) %>%
  summarise(Total = n()) %>%
  arrange(desc(Total)) %>%
  head()

#find the tailnum with the most occurence of late aircraft delay; N679AE
flight %>%
  filter(Origin == "ORD", LateAircraftDelay > 0) %>%
  group_by(TailNum) %>% 
  summarise(Total = n()) %>%
  arrange(desc(Total)) %>%
  head()

#From the filters, can see on 2 january 2006, the flight from MSP to ORD had a delay, which affected the subsequent flights
fdata <- flight %>%
  filter(TailNum == "N679AE") %>%
  select(Year, Month, DayofMonth, DepTime, CRSDepTime, ArrTime, CRSArrTime, DepDelay, ArrDelay, Origin, Dest) %>%
  arrange(Year,Month, DayofMonth, CRSDepTime) %>%
  filter(Year == "2006", Month == "1", DayofMonth == '2') %>%
  slice(1:3)

fdata

ggplot(fdata)+
  geom_line(aes(y=CRSArrTime,x= Origin,colour="CRSArrTime"),group=1 )+
  geom_point(aes(y=CRSArrTime,x= Origin,colour="CRSArrTime")) +
  geom_line(aes(y=ArrTime,x= Origin,colour="ArrTime"),group=1) +
  geom_point(aes(y=ArrTime,x= Origin,colour="ArrTime")) +
  labs(y ='Time')

```
### Question 5

```{r}
library(mlr3)
library(mlr3learners)
library(mlr3pipelines)
library(mlr3tuning)
library(skimr)
```

Create the column `Delayed`

```{r}
flight$Origin <- as_factor(flight$Origin)
flight %<>%
  mutate(Delayed = as_factor(ifelse(ArrDelay > 0 | DepDelay > 0, 1, 0)))

flight %<>% filter(!is.na(Delayed))

skim(flight)

delay_model <- flight %>%
  select(Year, Month, DayofMonth, DayOfWeek, CRSDepTime, CRSArrTime, ArrDelay, DepDelay, ADelay, DDelay, Delayed, Origin, DepInterval)

```

Create the training and test sets

```{r}
#get the training and test set
set.seed(1)

df <- sample_n(delay_model,100000)
n <- nrow(df)

train_set <- sample(n, round(0.7*n))
test_set <- setdiff(1:n, train_set)
```

### Set up the task 

```{r}
task <- TaskClassif$new('df', backend=df, target = 'Delayed')
task$select(c('Year', 'Month', 'DayofMonth', 'DayOfWeek', 'DepInterval'))
measure <- msr('classif.ce')
task
```
Convert factor variables into numerical values.

```{r}
fencoder <- po("encode", method = "treatment",
               affect_columns = selector_type("factor"))
ord_to_int <- po("colapply", applicator = as.integer,
                 affect_columns = selector_type("ordered"))
```

Some methods require tuning the hyperparameters.

```{r}
tuner <- tnr('grid_search')
terminator <- trm('evals', n_evals = 20)
```
## Different classification models

### Logistic regression

```{r}
learner_lr <- lrn("classif.log_reg")
gc_lr <- po('imputemean') %>>%
  po(learner_lr)
glrn_lr <- GraphLearner$new(gc_lr)

glrn_lr$train(task, row_ids = train_set)
glrn_lr$predict(task, row_ids = test_set)$score() 
```

### Gradient boosting

```{r}
library(xgboost)
learner_gb <- lrn("classif.xgboost")
gc_gb <- po('imputemean') %>>%
  fencoder %>>% ord_to_int %>>%
  po(learner_gb)
glrn_gb <- GraphLearner$new(gc_gb)

glrn_gb$train(task, row_ids = train_set)
```

```{r}
glrn_gb$predict(task, row_ids = test_set)$score() 
```

### Penalised logistic regression

```{r echo = TRUE, results = 'hide'}
learner_plr <- lrn('classif.glmnet') 
gc_plr <- po('scale') %>>% 
  fencoder %>>% ord_to_int %>>%
  po('imputemean') %>>%
  po(learner_plr)
glrn_plr <- GraphLearner$new(gc_plr)
tune_lambda <- ParamSet$new (list(
 ParamDbl$new('classif.glmnet.lambda', lower = 0.001, upper = 2)
))

at_plr <- AutoTuner$new(
  learner = glrn_plr,
  resampling = rsmp('cv', folds = 3),
  measure = measure,
  search_space = tune_lambda,
  terminator = terminator,
  tuner = tuner
)
at_plr$train(task, row_ids = train_set)
```

```{r}
at_plr$predict(task, row_ids = test_set)$score()
```

### Classification trees

```{r}
learner_tree <- lrn("classif.rpart")

gc_tree <- po('imputemean') %>>%
  po(learner_tree)
glrn_tree <- GraphLearner$new(gc_tree)

glrn_tree$train(task, row_ids = train_set)
glrn_tree$predict(task, row_ids = test_set)$score() 
```

### Random forests

```{r echo = TRUE, results = 'hide'}
learner_rf <- lrn('classif.ranger') 
learner_rf$param_set$values <- list(min.node.size = 4)
gc_rf <- po('scale') %>>%
  po('imputemean') %>>%
  po(learner_rf)
glrn_rf <- GraphLearner$new(gc_rf)
tune_ntrees <- ParamSet$new (list(
 ParamInt$new('classif.ranger.num.trees', lower = 50, upper = 600)
))
at_rf <- AutoTuner$new(
  learner = glrn_rf,
  resampling = rsmp('cv', folds = 3),
  measure = measure,
  search_space = tune_ntrees,
  terminator = terminator,
  tuner = tuner
)
at_rf$train(task, row_ids = train_set)
```

```{r}
at_rf$predict(task, row_ids = test_set)$score() 
```

### Support vector machines

```{r}
library(e1071)
learner_svm <- lrn("classif.svm")

gc_svm <- po('imputemean') %>>% 
  fencoder %>>% ord_to_int %>>%
  po(learner_svm)
glrn_svm <- GraphLearner$new(gc_svm)

glrn_svm$train(task, row_ids = train_set)
glrn_svm$predict(task, row_ids = test_set)$score() 
```

```{r results = 'hide'}
set.seed(1) # for reproducible results
# list of learners
lrn_list <- list(
  glrn_lr,
  glrn_gb,
  at_plr,
  glrn_tree,
  at_rf,
  glrn_svm
)
# set the benchmark design and run the comparisons
bm_design <- benchmark_grid(task = task, resamplings = rsmp('cv', folds = 3), learners = lrn_list)
bmr <- benchmark(bm_design, store_models = TRUE)
```

```{r}
library(mlr3viz)
library(ggplot2)
autoplot(bmr) + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
bmr$aggregate(measure)
```
